Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-15 Paper/Paper.tex
more changes - captions - some text improvements

% a short form should be given in case it is too long for the running head
\titlerunning{Enhancing Accelerometer-based Activity Recognition}

% the name(s) of the author(s) follow(s) next
\author{Tobias Grosse-Puppendahl\inst{1} \and Eugen Berlin\inst{2} \and Marko Borazio\inst{2} }
%
\authorrunning{Enhancing Accelerometer-based Activity Recognition}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{
	Fraunhofer IGD, Fraunhoferstr. 5, 64283 Darmstadt, Germany\\
	\email{tobias.grosse-puppendahl@igd.fraunhofer.de}
	\and
	Technische Universit\"at Darmstadt, Hochschulstr. 10, 64289 Darmstadt, Germany\\
	\email{\{berlin,borazio\}@ess.tu-darmstadt.de}
}

@@

\section{Introduction}
Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors attached to the body to identify the activities performed. However, it is a great challenge to identify fine-grained activities just by using this single modality. In this paper, we use an accelerometer-based activity recognition system, the wrist-worn activity logger called HedgeHog\footnote{\url{http://www.ess.tu-darmstadt.de/hedgehog}} and investigate the possibility of integrating a capacitive proximity sensor unobtrusively into the wristband to obtain higher recognition rates for several activities in daily life.

@@

Berlin et al. \cite{berlin_laerhoven_tei_2010} have augmented an accelerometer based sensor node with an RFID bracelet to be able to detect objects a human is interacting with. Thus, activities can use this additional modality

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in [wimmer2006] who presented a toolkit for capacitive proximity sensing.

@@

\section{Hardware}
\subsection{HedgeHog}
The HedgeHog sensor is a custom designed and built wearable data logger aiming at long-term deployments in activity recognition scenarios. Due to its small form-factor (30x25x20mm) and weight, this sensor, being worn at the wrist, is an unobtrusive way to record relevant motion data.

The HedgeHog sensor is built around the low-power Microchip microcontroller (PIC18F46J50) featuring an accelerometer sensor (ADXL345) to capture human motion, light and ambient temperature sensors and a microSD flash card for locally storing the sensor data. The sensor is powered by a $200mAh$ lithium polymer battery, which allows for two weeks of continuous recording on a single battery charge. A plastic case is used to nicely package and protect the sensor to be worn at the wrist (as shown in Fig. \ref{fig:sensornode}).

@@

The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor, which is described in detail in the next section.

@@

pseudocode:
    sample the accelerometer sensor

start frequency counter

go to sleep for 9.5 ms

wake-up (timer interrupt)

stop frequency counter

read frequency count

@@

@@

\begin{figure}
	\centering
		\includegraphics[trim=2cm 0.5cm 2cm 0.5cm,width=1.00\textwidth]{Images/drinking.pdf}
	\caption{Drinking activity}
	\label{fig:drinking}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[trim=2cm 0.5cm 2cm 0.5cm,width=1.00\textwidth]{Images/making-bread.pdf}
	\caption{Preparing bread}
	\label{fig:making-bread}
\end{figure}

>>>>

% a short form should be given in case it is too long for the running head
\titlerunning{Enhancing Accelerometer-based Activity Recognition}
% (feature abused for this document to repeat the title also on left hand pages)
\authorrunning{Enhancing Accelerometer-based Activity Recognition}

% the name(s) of the author(s) follow(s) next
% \author{Tobias Grosse-Puppendahl\inst{1} \and Eugen Berlin\inst{2} \and Marko Borazio\inst{2} }
\author{Author 1 \and Author 2 \and Author 3 }

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
% \institute{
% 	Fraunhofer IGD, Fraunhoferstr. 5, 64283 Darmstadt, Germany\\
% 	\email{tobias.grosse-puppendahl@igd.fraunhofer.de}
% 	\and
% 	Technische Universit\"at Darmstadt, Hochschulstr. 10, 64289 Darmstadt, Germany\\
% 	\email{\{berlin,borazio\}@ess.tu-darmstadt.de}
% }

\institute{
	Anonymized for review \\
	\email{ananymous@insitution.edu}
	\and
	Anonymized for review \\
	\email{ananymous@insitution.edu}
}

@@

\section{Introduction}
Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors attached to the body to identify the activities performed. However, it is a great challenge to identify fine-grained activities just by using this single modality. In this paper, we have obtained and use a wrist-worn activity data logger\footnote{HedgeHog Activity Logger -- \url{http://www.ess.tu-darmstadt.de/hedgehog}} and investigate the possibility of integrating a capacitive proximity sensor unobtrusively into the wristband to obtain higher recognition rates for several activities in daily life.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The field of activity recognition relying on wearable sensors mostly mostly considers inertial data from the participants body to infer performed activities, such as in the works of [bla,foo]. The acceleration data is often augmented with data from sensors such as gyroscopes, magnetometers, ambient light or temperature, aiming at extracting more motion and environmental context of the user. [references?!]

The works of \cite{intel_bracelet} and ... show that detecting objects the user touches or uses throughout the day can be very helpful for the activity recognition, including activities of daily living (maja(?)), maintenance tasks (?), gardening activities, etc. For example, Berlin et al. \cite{berlin_laerhoven_tei_2010} have augmented a wrist-worn accelerometer sensor node with an RFID bracelet to be able to detect objects a human is interacting with.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{wimmer2006} who presented a toolkit for capacitive proximity sensing.

@@

\section{Hardware}
This section presents the two components of the required hardware, the wrist-worn activity data logger tailored to capture acceleration data, and the capacitive proximity sensor used for distance measurements.

\subsection{Activity Data Logger}
The HedgeHog sensor is a custom designed and built wearable data logger aiming at long-term deployments in activity recognition scenarios. Due to its small form-factor (30x25x20mm) and weight, this sensor, being worn at the wrist, is an unobtrusive way to record relevant motion data.

The sensor node itself is built around the low-power Microchip microcontroller (PIC18F46J50) featuring an accelerometer sensor (ADXL345) to capture human motion, light and ambient temperature sensors and a microSD flash card for locally storing the sensor data. The sensor is powered by a $200mAh$ lithium polymer battery, which allows for two weeks of continuous recording on a single battery charge. A USB port is used to configure the sensor (e.g. setting the sensitivity of the accelerometer), to access the stored sensor data, and to recharge the battery. A plastic case nicely packages and protects the sensor to be worn at the wrist (as shown in Fig. \ref{fig:sensornode}).

@@

Since the main microcontroller on the sensor node has unoccupied pins, it can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor, which is described in detail in the next section.

@@

\begin{verbatim}
pseudocode:
  sample the accelerometer sensor
  start frequency counter
  go to sleep for 9.5 ms
  wake-up (timer interrupt)
  stop frequency counter
  read frequency count
\end{verbatim}

@@

This section presents the experimental setup including the activities and the participants, as well as the findings that were obtained during the evaluation.

@@

\begin{figure}
	\centering
		\includegraphics[trim=1.9cm 0.5cm 1.8cm 0.5cm,clip,width=\textwidth]{Images/drinking.pdf}
	\caption{An example of the ``drinking'' activity with raw 3D acceleration data (lower plot) and proximity sensor data (upper plot). The participant took five nips from the glass, while leaving the hands lying on the table in between. These motions can be easily detected in the acceleration as well as proximity data.}
	\label{fig:drinking}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[trim=1.9cm 0.5cm 1.8cm 0.5cm,clip,width=\textwidth]{Images/making-bread.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a toast. Note, that the acceleration data does not seem to have any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in ``unmittelbarer nähe''.}
	\label{fig:making-bread}
\end{figure}
